---
title: "Lab 2 (3.2.2026)"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# Overview of Lab 2

Welcome to R Lab 2! In this lab, we will do two different exercises:

1)  Explore how to analyze and visualize network data with the package `tidygraph`
2)  Investigate how much homophily there is in our dicussion networks using the survey data we collected in class

Each exercise will have 2-3 questions. Please complete these questions and upload the .qmd file to the corresponding lab assingment Canvas to receive full credit for the lab.

# Exercise 1: Analyze network data

We will use the Florentine Family Marriage dataset as a running example for network analysis. Before we start, we need to install and load required packages.

`tidygraph` includes tools for network analysis. `ggraph` is used for visualizing networks. `networkdata` contains network datasets, including the Florentine marriage network we will use.

Recall, that to install a package, we need the function `install.packages()`.

```{r}
# install packages

# install.packages("tidygraph")
# install.packages("ggraph")
# devtools::install_github("schochastics/networkdata")
```

Next, we'll need to library the packages with the function `library()`.

```{r}
# load packages
library(tidygraph)
library(networkdata)
library(ggraph)
library(tidyverse)
```

Now we load the Florentine Family Marriage dataset `flo_marriage` into the current R session with the function `data()`. This data is from the `networkdata` pacakge.

The data originate from research on social relations among Renaissance Florentine families collected by John Padgett from historical documents. Breiger and Pattison (1986) later used this subset of the data in their analysis of social roles.

```{r}
data("flo_marriage")
```

The `flo_marriage` is stored as a network igraph object. To use `tidygraph`, we need to convert this network object into a tidygraph object using `as_tbl_graph` so we can analyze it using tidy tools. We assign a name `flo_tbl` to the converted data frame.

```{r}
flo_tbl <- as_tbl_graph(flo_marriage)
flo_tbl
```

The dataset contains two components, a node table and an edge table. Each node represents a Florentine family in the early 15th century, and each edge represents a marriage tie between two families.

There are four columns (variables) in the node table:

-   `name`: the names of Florentine Families.
-   `wealth`: each family's net wealth in 1427 (reported assets minus deductions, including business loans) from 1427 catasto (in thousands of lira).
-   `#priors`: the number of priorates (seats on the civic council) held between 1282-1344, reflecting the family influence.
-   `ties`: the total number of business or marriage ties in the total dataset of 116 families.

There are two columns (variables) in the edge table:

-   `from`: the node ID that the edge starts from
-   `to`: the node ID that the edge connects to

The edge table gives us the information on which nodes are connected as marriage ties. Node IDs are used instead of family names because numerical IDs are more efficient for computation purpose.

You can find more information about the dataset from the study Breiger & Pattison (1986) here:

https://doi.org/10.1016/0378-8733(86)90006-7

Breiger R. and Pattison P. (1986). Cumulated social roles: The duality of persons and their algebras. Social Networks, 8, 215-256.

Overall, we can learn from the output that:

-   The network has 16 families (nodes) and 20 marriage ties (edges).

-   It is undirected (marriage is mutual).

If you want to check explicitly the number of nodes and edges, use the following two functions from the `tidygraph` package.

```{r}
# Number of nodes
flo_tbl %>% 
  activate(nodes) %>%
  as_tibble() %>%
  summarise(n_nodes = n())

# Number of edges
flo_tbl %>% 
  activate(edges) %>% 
  as_tibble() %>%
  tally(name = "n_edges")
```

In the code chunk above, let's focus on the nodes-related lines first:

`%>%` is the pipe operator that lets us chain operations together without creating intermediate objects. It works by taking the object on the left-hand side and passing it as the first argument to the next function.

`activate()` tells `tidygraph` to apply subsequent manipulations to the nodes, since the tbl_graph objects hold two separate, linked tables (node and edges).

`as_tibble()` converts the tbl_graph object into a regular table so that standard data-frame functions like summarise() can work. A tbl_graph is not a regular data frame.

`summarise()` collapses a table into a summary and produces one row with summary statistics.

`n()` counts the number of rows in the current table. Since we activated nodes, it counts the number of nodes.

`n_nodes = n()` creates a new column called n_nodes and stores the count in that column.

So `summarise(n_nodes = n())` simply counts how many rows are in the node table and call that number n_nodes.

The same logics apply to the edges-related lines.

Instead of `summarise(n_nodes = n())`, we can also use `tally(name = "n_edges")` to count the number of rows in the edge table and name the number to "n_edges".

## The Medici Family

Now that we have a basic understanding of the network, let's do something fun. The Medici family was a powerful banking dynasty in Florence that produced renowned patrons of the arts, four popes, and two French queens, leaving a lasting impact on European economy, culture, and politics.

But here's our question: *Can we observe the Medici family's influence from the family marriage network structure?*

Let's visualize the network first using the `ggraph` package.

```{r}
set.seed(8287) # fix randomness so that our results can be reproduced

ggraph(flo_tbl, layout = "stress") +
  geom_edge_link(alpha = 0.4) +
  geom_node_point(size = 3) +
  geom_node_text(aes(label = name), repel = TRUE) +
  theme_void()
```

In the code chunk above,

`set.seed()`: fix the starting point of R's random number generator. Without setting a seed (e.g., 8287), layout algorithms in `ggraph` that rely on randomness may produce slightly different network visualizations each time the code is executed.

`ggraph(flo_tbl, layout = "stress")`: a function used to visualize networks. `flo_tbl` is the network object being plotted, and `layout = "stress"` specifies a force-directed layout algorithm that positions nodes so that connected nodes appear closer together in the visualization.

`+`: the syntax allows us to add layers to the plot.

`geom_edge_link(alpha = 0.4)`: a function used to draw the edges of the network, so that each row in the edge table becomes a connection (tie) in the visualization. The argument `alpha = 0.4` controls the transparency of the edge lines, where 0 is fully transparent and 1 is fully opaque. Setting it to 0.4 makes the edges lighter, which helps reduce visual clutter and makes the nodes easier to distinguish.

`geom_node_point(size = 3)`: a function used to draw the nodes as points (circles), so that each row in the node table corresponds to one node in the visualization. The argument `size = 3` controls how large the nodes appear in the plot.

`geom_node_text(aes(label = name), repel = TRUE)`: a function used to add text labels to nodes. `aes()` stands for aesthetic mapping, and it tells R to use the variable name from the node table as the label. `repel = TRUE` prevents labels from overlapping and makes the plot more readable.

`theme_void()`: a function that removes axes, ticks, and background grid to make the plot cleaner.

#### Exercise Question 1.1

Discuss what you see from the network graph. Does the Medici family appear central in the structure?

### Centrality Measures

Now we move to measurements for centrality. We will calculate two types of centrality: degree and betweenness centrality.

```{r}
flo_tbl %>%
  activate(nodes) %>%
  mutate(degree = centrality_degree(),
         betweenness = centrality_betweenness()) %>%
  as_tibble() %>%
  select(name, degree, betweenness)
```

In the code chunk above, we first tell R to apply subsequent manipulations to the nodes using `activate(nodes)`. Then we have the following functions:

`mutate()`: add new columns to the active table (which is currently the node table).

`centrality_degree()`: calculate the degree centrality (the number of connections) for each node. The result is stored in a new column called `degree`.

`centrality_betweenness`: calculate the betweenness centrality (how often a node lies on the shortest path between other nodes) for each node. The result is stored in a new column called `betweenness`.

With `as_tibble()`, we again convert the tbl_graph object into a regular table.

`select()`: chooses only the columns we want to display. Here, we want to display only the family names, degree centrality, and betweenness centrality values.

#### Exercise Question 1.2

What does each centrality measure reveal about the Medici family's influence, and what does this imply about its power in the network?

Now let's combine the two centrality measures into one visualization.

```{r}
set.seed(8287)

flo_tbl %>% 
  activate(nodes) %>% 
  mutate(
    degree = centrality_degree(),
    betweenness = centrality_betweenness()
  ) %>% 
  ggraph(layout = "stress") + 
  geom_edge_link0(alpha = 0.4) +
  geom_node_point(
    aes(size = degree, fill = betweenness),
    shape = 21
  ) +
  geom_node_text(
    aes(label = name), repel = TRUE, size = 3.5) +
  scale_fill_viridis_c( option = "cividis", name = "Betweenness") +
  scale_size(range = c(5, 10), name = "Degree") +
  theme_graph() +
  labs(title = "Florentine Marriage Network")
```

In the code chunk above, we first set the seed to 8287 to ensure reproducibility. We then tell `tidygraph` to apply subsequent operations to the node table using `activate("nodes")`, and add the computed degree and betweenness centrality using `mutate()`.

Next, we use `ggraph()` to create the network visualization and add layers step by step.

`geom_edge_link0(alpha = 0.4)`: This draws the edges (connections) between nodes. As in the previous visualization, we set `alpha = 0.4` to make the edges slightly transparent, which reduces visual clutter and makes the nodes easier to see.

`geom_node_point()`: This draws the nodes as points. Inside `aes()` (aesthetic mapping), we specify `size = degree` and `fill = betweenness`, meaning that node size represents degree centrality and node color represents betweenness centrality. This allows us to compare the two measures visually. The argument `shape = 21` enables both fill color and border color for the nodes, which is necessary when mapping fill aesthetics.

`geom_node_text()`: This adds labels to each node. We specify `label = name` to display each family's name, use `repel = TRUE` to prevent text overlap, and control the text size with `size = 3.5`.

`scale_fill_viridis_c(option = "cividis", name = "Betweenness")`: This defines the color scale for betweenness centrality. We use the colorblind-friendly "cividis" palette and label the legend as "Betweenness."

`scale_size(range = c(5, 10), name = "Degree")`: This controls the visual size range of the nodes, scaling them between 5 and 10, and labels the legend as "Degree."

`theme_graph()`: This removes axes and background elements to create a cleaner network visualization.

`labs(title = "Florentine Marriage Network")`: This allow us to add or modify labels in a plot. Here, we add a title for the plot with `title = "Florentine Marriage Network"`.

#### Exercise Question 1.3

What would happen if we remove Medici from the network?

# Exercise 2: Homophily Survey

# Bonus: Infectious Diseases in Networks (Optional)

If you are curious about how network structure shapes disease spread, you can explore a Shiny app called SimNetwork developed by Alison Wu:

https://github.com/alisonswu/shiny-SimNetwork

This interactive app allows you to: \\

-   Select different network structures

-   Adjust disease parameters (e.g., transmission rate, recovery rate)

-   Visualize SIR-like infectious disease dynamics on networks

You may run this on your own time if you'd like to experiment with dynamic processes on networks. To launch the app locally, you will need to run this code:

```{r}
install.packages("shiny")
library(shiny)

runGitHub("shiny-SimNetwork", "alisonswu")
```

`runGitHub()` is a function from the `shiny` package that allows us to run a Shiny app directly from a GitHub repository.

This app is divided into three main steps:

1.  Choose a network structure

Under "Step 1: Choose Network", you select

-   A network type among options `Scale Free`, `Small World`, `Scale Free`, and `Lattice` from the dropdown menu.

-   `Population size`, which determines how many individuals (nodes) are included in the network.

2.  Choose disease parameters

Under "Step 2: Choose disease dynamic", you can alter the following parameters:

-   `Number of Initial Infections`: the number of individuals who are infected at time t = 1.

-   `Infection Probability`: the probability (ranging from 0 to 1) that the infection transmits across a connection.

-   `Infection Duration`: how long an individual remains infected before transitioning to the recovered state.

3.  Run the simulation

Click the `simulate` button to begin the simulation.

The output is displayed in two main tabs:

-   Network visualization

This shows nodes (individuals), edges (connections), color-coded health status including susceptible (green), infected (red), recovered (yellow).

You can move the time slider to observe how the infection spreads step by step. You can also click `start/pause animation` (Step 4) to automatically play or pause the dynamic process.

-   SIR plots (line plot and stacked area plot)

They show the population-level dynamics over time: the percentage of Susceptible (S), the percentage of Infected (I), the percentage of Recovered (R). The line plot emphasizes trends and infection peaks, while the stacked area plot shows how the population composition changes over time.

## Task 1

Run the simulation for different network structures while keeping all disease parameters at their default (unchanged) values.

#### Bonus Question 1

Which network structure produced the largest outbreak? What features of that network's structure might make disease spread more easily?

## Task 2

Choose two disease parameters and experiment with them. Vary one parameter at a time while keeping the network type and all other settings fixed. After each change, run the simulation and observe.

#### Bonus Question 2

Does changing the disease dynamics alter which network structure is most vulnerable to large outbreaks?
