---
title: "Lab 2"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# Overview of lab 2

Welcome to R Lab 2! In this lab, we will do two different exercises:

1)  Explore how to analyze and visualize network data with the package `tidygraph`
2)  Investigate how much homophily there is in our discussion networks using the survey data we collected in class

Each exercise will have 2 questions. Please complete these questions and upload the .Rmd file to the corresponding lab assingment Canvas to receive full credit for the lab.

## Exercise 1: Analyze network data

We will use the Florentine Family Marriage dataset as a running example for network analysis. Before we start, we need to install and load required packages.

`tidygraph` includes tools for network analysis. `ggraph` is used for visualizing networks. `networkdata` contains network datasets, including the Florentine marriage network we will use.

Recall, that to library a package, we need to first install it.

```{r}
# install packages

# install.packages("tidygraph")
# install.packages("ggraph")
# devtools::install_github("schochastics/networkdata")
```

Next, we'll need to library (load) the packages. We need to install a package only once, but library it (load it) every time we start a new R session.

```{r}
# load packages
library(tidygraph)
library(networkdata)
library(ggraph)
library(tidyverse)
```

Now we load the Florentine Family marriage network. This data is from the `networkdata` package.

```{r}
data("flo_marriage")
```

The `flo_marriage` is stored as a network (igraph) object. To work with it using tidygraph, we convert it into a `tbl_graph`, which allows us to manipulate nodes and edges using tidyverse syntax.

```{r}
flo_tbl <- as_tbl_graph(flo_marriage)
flo_tbl
```

Each node represents a Florentine family in the early 15th century, and each edge represents a marriage tie between two families. The output tells us:

-   The network has 16 families (nodes) and 20 marriage ties (edges).

-   It is undirected (marriage is mutual).

If you want to check explicitly the number of nodes and edges, use the following two functions from the `igraph` package embedded in the `tidygraph` package.

```{r}
# Number of nodes
flo_tbl %>% 
  activate(nodes) %>% 
  as_tibble() %>%
  count()

# Number of edges
flo_tbl %>% 
  activate(nodes) %>% 
  as_tibble() %>%
  count()
```

## The Medici Family

Now that we have a basic understanding of the network, let's do something fun. The Medici family was a powerful banking dynasty in Florence that produced renowned patrons of the arts, four popes, and two French queens, leaving a lasting impact on European economy, culture, and politics.

But here's our question: *Can we observe the Medici family's influence from the family marriage network structure?*

Let's visualize the network first.

```{r}
## fix randomness so our graphs all look the same 
set.seed(1) 

## create a visualization of the network 
ggraph(flo_tbl, layout = "stress") +
  geom_edge_link(alpha = 0.4) +
  geom_node_point(size = 3) +
  geom_node_text(aes(label = name), repel = TRUE) +
  theme_void()
```

**Question: Discuss what you see from the network graph. Does the Medici family appear central in the structure?**

### Centrality Measures

Now we move to measurements for centrality. We will calculate two types of centrality: degree and betweenness centrality.

```{r}
flo_tbl %>%
  activate(nodes) %>%
  mutate(degree = centrality_degree(),
         betweenness = centrality_betweenness()) %>%
  as_tibble() %>%
  select(name, degree, betweenness)
```

**Question: What does each centrality measure reveal about the Medici family's influence, and what does this imply about its power in the network?**

Now let's combine the two centrality measures into one visualization.

```{r}
set.seed(8287)

flo_tbl %>% 
  activate("nodes") %>% 
  mutate(
    degree = centrality_degree(),
    betweenness = centrality_betweenness()
  ) %>% 
  ggraph(layout = "stress", bbox = 10) + 
  geom_edge_link0(edge_color = "black") +
  geom_node_point(
    aes(size = degree, fill = betweenness),
    shape = 21
  ) +
  geom_node_text(
    aes(label = name), repel = TRUE, size = 3.5) +
  scale_fill_viridis_c( option = "cividis", name = "Betweenness") +
  scale_size(range = c(5, 10), name = "Degree") +
  theme_graph() +
  labs(title = "Florentine Marriage Network")
```

**Question: What would happen if we remove Medici from the network?**

# Exercise 2: Homophily Survey

# Bonus: Infectious Diseases in Networks (Optional)

If you are curious about how network structure shapes disease spread, you can explore a Shiny app called SimNetwork developed by Alison Wu:

https://github.com/alisonswu/shiny-SimNetwork

This interactive app allows you to:

-   Select different network structures

-   Adjust disease parameters (e.g., transmission rate, recovery rate)

-   Visualize SIR-like infectious disease dynamics on networks

You may run this on your own time if you’d like to experiment with dynamic processes on networks. To launch the app locally, you will need to run this code:

```{r}
install.packages("shiny")
library(shiny)
runGitHub("shiny-SimNetwork", "alisonswu")
```

## Exercises

**3.1**: Run the simulation using different network structures. Keep all disease settings the same. Which network structure produced the largest outbreak? What features of that network's structure might make disease spread more easily?

**3.2**: Choose two disease parameters and experiment with them. Change one parameter at a time while keeping the network type set to “small world” and all other settings the same. After each change, run the simulation and observe the results. Which parameters, when increased, make it more likely that the entire population becomes infected? List the parameters and briefly explain why increasing them leads to a larger outbreak.
