---
title: "Social Networks (Soc 321k)"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
# load packages
library(tidyverse)

# set theme for plots 
theme_set(theme_bw(base_size = 17))
```

# Overview of Lab 3

Welcome to R Lab 3! In this lab, we will complete two different exercises:

1)  Explore how network structure shapes disease spread using simulated data in the Shiny app (SimNetwork).

2)  Investigate patterns of AI use using the survey data collected in class.

Each exercise will have a few questions for you to complete. Please complete these questions and upload the .qmd file to the corresponding lab assignment to Canvas to receive full credit for the lab.

# Exercise 1: SIR Models on Networks

In Exercise 1, we are interested in how network structure shapes disease spread. We explore a Shiny app demo'd in class called SimNetwork developed by Alison Wu:

https://github.com/alisonswu/shiny-SimNetwork

This interactive app allows you to:

-   Select different network structures

-   Adjust disease parameters (e.g., transmission rate, recovery rate)

-   Visualize SIR-like infectious disease dynamics on networks

To launch the app locally, you will need to run this code:

```{r}
## uncomment lines below to install and run the shiny app
# install.packages("shiny")
# library(shiny)
# 
# runGitHub("shiny-SimNetwork", "alisonswu")
```

`runGitHub()` is a function from the `shiny` package that allows us to run a Shiny app directly from a GitHub repository.

This app is divided into three main steps:

1.  Choose a network structure

Under "Step 1: Choose Network", you select

-   A network type among options `Scale Free`, `Small World`, `Binary Trees`, and `Lattice` from the dropdown menu.

-   `Population size`, which determines how many individuals (nodes) are included in the network.

2.  Choose disease parameters

Under "Step 2: Choose disease dynamic", you can alter the following parameters:

-   `Number of Initial Infections`: the number of individuals who are infected at time t = 1.

-   `Infection Probability`: the probability (ranging from 0 to 1) that the infection transmits across a connection.

-   `Infection Duration`: how long an individual remains infected before transitioning to the recovered state.

3.  Run the simulation

Click the `simulate` button to begin the simulation.

The output is displayed in two main tabs:

-   Network visualization

This shows nodes (individuals), edges (connections), color-coded health status including susceptible (green), infected (red), recovered (yellow).

You can move the time slider to observe how the infection spreads step by step. You can also click `start/pause animation` (Step 4) to automatically play or pause the dynamic process.

-   SIR plots (line plot and stacked area plot)

They show the population-level dynamics over time: the percentage of Susceptible (S), the percentage of Infected (I), the percentage of Recovered (R). The line plot emphasizes trends and infection peaks, while the stacked area plot shows how the population composition changes over time.

## Exercises

**1.1**: Run the simulation using different network structures. Set the infection probability to p = 0.1. Keep all disease settings the same. Which network structure produced the largest outbreak? What features of that network's structure might make disease spread more easily?

*Small world networks produced the largest outbreak. These networks combine dense local clustering with a few long-range "shortcut" ties, which dramatically shortens the average path between any two nodes. That combination lets infection spread efficiently within neighborhoods while also jumping quickly across the network.*

**1.2**: Change one parameter at a time while keeping the network type set to "small world" and all other settings the same. After each change, run the simulation and observe the results. Which of the three parameters, when increased or decreased, make it more likely that the entire population becomes infected? List each parameters and briefly explain why increasing/decreasing them leads to a larger outbreak.

*Decreasing population size makes it less likely that the entire population becomes infected, because smaller networks saturate quickly and there are fewer nodes for the infection to miss. Increasing the number of initial infections makes a full outbreak more likely, since multiple seeds launch separate transmission chains simultaneously and are less likely to all die out by chance early on. Increasing the infection probability also makes a full outbreak more likely, because each contact between an infected and susceptible node is more likely to successfully transmit, raising the effective reproduction number above the threshold needed for widespread spread.*

# Exercise 2: Homophily Survey

Now we'll turn to analyzing data from our homophily survey. In this survey, we collected information about an ego and five of their discussion partners (alters). All nonsensical responses and missing data have been dropped. We'll read this file in from where it is stored (on Github) using the `read_csv()` function.

The data is formatted in `long` format, where each row represents a different alter. This means that we will have 5 rows for each ego.

```{r}
## read in survey responses 
responses <- read_csv("https://raw.githubusercontent.com/caseybreen/networks_course/main/data/homophily_survey.csv")

## print out survey responses 
responses
```

Let's look at the columns:

`ego_id`: A unique identifier for each survey respondent (the ego in the ego network). Each respondent will appear five times in the dataset, once for each person they listed.

`date`: The date the survey response was submitted.

`ego_gender`: The gender of the respondent completing the survey.

`ego_age`: The age of the respondent.

`ego_classyear`: The respondent's class year (Freshman, Sophomore, Junior, or Senior).

`ego_origin`: Where the respondent is from (e.g., Austin, Rest of Texas, Rest of the United States, or Rest of the world).

`ego_ai`: Whether the respondent reports having used AI (e.g., ChatGPT) on coursework in a way that violated course rules in the past 12 months.

`alter_id`: A numeric identifier for each discussion partner listed by the respondent. Each respondent could list up to five people, so this variable ranges from 1 to 5.

`alter_classyear`: The class year of the discussion partner listed by the respondent (if they are a student).

`alter_gender`: The gender of the discussion partner listed by the respondent.

`alter_origin`: Where the discussion partner listed by the respondent is from.

`alter_age`: The age of the discussion partner listed by the respondent.

`alter_ai`: Whether the respondent believes that the discussion partner person has used AI (e.g., ChatGPT) on coursework in a way that violated course rules in the past 12 months.

## Warm-Up Exercises

Recall that in Lab 2, we did some simple explorations of this dataset. We examined the age and gender distributions of egos and alters, and we looked at homophily in characteristics, such as age, gender, and year in college, between egos and their discussion partners.

**a)** Create a histogram plot of the ages of our discussion partners in the following code chunk. Use the `ggplot()` and `geom_histogram` commands.

```{r}
## make histogram 
responses %>% 
  ggplot(aes(x = alter_age)) + 
  geom_histogram()
```

**b)** Examine homophily with respect to gender in the discussion networks by completing and running the code below. Do you see evidence of gender homophily?

```{r}
## tabulate discussion partners gender 
 gender_tabulated <- responses %>% 
   count(ego_gender, alter_gender)

## plot discussion partners gender 
gender_tabulated %>%
  ggplot(aes(x = alter_gender, y = ego_gender, fill = n)) +
  geom_tile() +
  geom_text(aes(label = n), size = 7) +
  scale_fill_viridis_c(option = "cividis") +
  labs(
    x = "Alter Gender",
    y = "Ego Gender",
    fill = "Count"
  )
```

## Unauthorized AI use 

We now shift our focus to a different topic: reported unauthorized use of AI in the past 12 months.

Before we begin, let's briefly recall the **network scale-up method**:

A technique for estimating the size of hard-to-reach ("hidden") populations, such as drug injectors and sex workers, by leveraging sampled social network data. It works by asking survey respondents how many people they know in a target group, then using that ratio to estimate the total population size.

We apply a similar idea in this context. Because unauthorized AI use may be stigmatized, students may underreport their own behavior when asked directly. However, they may be more willing to report such behavior among their peers. To explore this, we compare two measures of unauthorized AI use:

1.  Self-reported unauthorized AI use among students in the class (egos)
2.  Perceived unauthorized AI use among their UT student discussion partners (alters)

```{r}
## calculate ego responses 
ego_ai_response <- responses %>%
  distinct(ego_id, ego_ai) %>%
  summarise(
    count = n(),
    count_use_ai = sum(ego_ai == "Yes"),
    prop_use_ai = mean(ego_ai == "Yes"),
    group = "egos"
  )

## print out responses 
ego_ai_response
```

In the above code, we uses the pipe `%>%` to pass the `responses` dataframe into the next function.

Recall that each ego appears five times in the data (once for each discussion partner). To avoid counting the same ego more than once, we first use:

`distinct(ego_id, ego_ai)`: This keeps only one row per ego based on `ego_id`. If the same `ego_id` appears multiple times, duplicates are removed and only the first occurrence is kept. After this step, the dataset contains one row per ego, that is, 15 rows in totals since there are 15 egos.

Next, we summarize the data:

`summarise()`: This function reduces the dataset to a single row containing summary statistics that we define.

Inside `summarise()`:

-   `count = n()`: Counts the number of rows at this stage, which now corresponds to the number of unique egos.

-   `count_use_ai = sum(ego_ai == "Yes")`: Counts how many egos reported having used AI (e.g., ChatGPT) on coursework in a way that violated course rules in the past 12 months.

    -   The expression `ego_ai == "Yes"` returns TRUE for egos who answered "Yes" and FALSE otherwise.
    -   In R, TRUE is treated as 1 and FALSE as 0, so sum() effectively counts the number of "Yes" responses.

-   `prop_use_ai = mean(ego_ai == "Yes")`: Calculates the proportion who reported unauthorized AI use. Taking the mean gives the proportion of "Yes" responses. The output is assigned to `prop_use_ai`.

-   `group = "egos"` creates a new column specifying the group we are studying is egos.

Update the code below to perform a similar calculation, but for alters. Assign the `group` value of "alters" instead of "ego."

```{r}
## calculate alter response 
alter_ai_response <- responses %>%
  filter(alter_ai != "Not a UT student") %>% ## only include UT students
  summarise(    
    count = n(),
    count_use_ai = sum(alter_ai == "Yes"),
    prop_use_ai = mean(alter_ai == "Yes"),
    group = "alters"
  ) 

## print out responses 
alter_ai_response
```

The code above performs a similar calculation, but for alters.

Here, we do not use `distinct()` because each row already represents a unique alter.

`filter(alter_ai != "Not a UT student")`: removes alters who are not UT students from the sample.

Next, we summarize the data almost parallel to the above.

### Exercises

**2.1**: Fill in the blanks in the code below to make a plot of the rates of unauthorized AI use among students in the class and among their UT student alters. Make the x-axis the group (ego vs. alters), the y-axis the percent using ai, and the color (fill argument) the group.

```{r}
## create one dataframe for plotting 
ego_and_alter_ai_response <- bind_rows(alter_ai_response, ego_ai_response)

## plot the result
ego_and_alter_ai_response %>% 
  ggplot(aes(x = group, y = prop_use_ai, fill = group)) + 
  geom_col() + 
  scale_y_continuous(labels = scales::percent_format()) 
```

**2.2**: Interpret. Which rate is higher? Why might students being surveyed report higher/lower rates for their peers than for themselves?

*The rate is higher for reports about alters than for self-reports. This likely reflects social desirability bias: students may be reluctant to admit in a survey that they personally used AI in ways that violated course rules, even when the survey is anonymous, because the behavior is stigmatized and sanctionable. They face no such reputational cost when reporting on their peers' behavior, so those estimates are probably closer to the true rate.*

**2.3**: How do our results illustrate the value of the network scale-up method? For something extremely stigmatized, such as injecting drugs, do you think the gap between self-reports and peer-reports would be higher or lower?

*Our results illustrate the core logic of the network scale-up method: when a behavior is stigmatized, asking people about their social contacts can yield more accurate estimates of its prevalence than asking them about themselves. By leveraging what respondents observe in their networks, we can get around the underreporting that direct self-reports produce. For something as stigmatized as injecting drug use, we would expect the gap between self-reports and peer-reports to be even larger, because the social and legal consequences of admitting to injecting drugs are far more severe than admitting to unauthorized AI use. This is exactly why the network scale-up method was originally developed for: estimating the size of hidden populations like people who inject drugs and sex workers.*
